From noise to knowledge: how randomness generates novel phenomena and reveals information

### 2018-03-02 paper/paper.Rmd
intro!

<!--
Directions:

We hope that these papers will combine a mix of synthetic reviews of the current state of an area, as well as insights into how the field has developed during the last twenty years, discussion of current important questions, and developments and insights that would significantly advance the field. Although certainly not necessary, if there are any past Ecology Letters papers that have had a profound impact on the development of an area – be it beneficial or not – a brief discussion of that impact would be appropriate. We also hope that articles will be written in a thoughtful style, but also provocative if appropriate.

@@

<!-- FIXME look at Coulson, Bjornstead, Ovaskeinen and others and see how they define scope to focus on "population ecology" -->

@@

The gap between theory-focused specialists and more empirically minded non-specialists
is particularly acute in the ecological literature on stochastic processes, making
a review of 
<!-- FIXME -->

As the literature on stochastic modeling in ecology is both extensive and often
technical, and so rather than attempt a comprehensive review, this synthesis
will focus on the simplest examples necessary to illustrate this broader trend
in the changing way we think about the role of noise.
To that end, I will focus on examples from simple population models, echoing
the motivation of @Bartlett1960 to focus on basic theoretical aspects common
across models. In so doing, we will ignore the equally rich literature of 
stochastic models in evolution [@Lenormand2009] and population genetics
[e.g. @Ewens2004; @Gillespie2004]. In an effort to make these themes accessible to more empirical
readers, I will rely on the simplest examples necessary, with only passing
mention of the rich literature extending these concepts to more realistic
models.  It is easy to critique this choice: indeed, one might summarize decades
of research in noting the twin observations that (1)  ecology is complex and (2)
that complexity is important. In the stochasticity literature alone, much work 
has been done to understand and extend fundamental results to more complex models,
including the importance of age and stage structure [e.g. @Cohen1979; @Tuljapurkar1989;
@Caswell2009; @Metcalf2015; @Vindenes2017b], spatial structure [e.g. @Durrett1994; @Dieckmann2000;
@Kerr2002; @Schreiber2010], Individual heterogeneity [e.g. @Vindenes2015; @Hart2016]
or the coexistence of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Schreiber2017]; 
correlated [@Ripa; @Ripa1999; @Petchey, @Schreiber2010; @Lee2017; @Spanio2017] or periodic
[@Coulson2001; @Bjornstad2001; @Keeling2001] structure in environmental noise, or
the interaction of ecological and evolutionary processes [@Saether1997; @Ozgul2009;
@Lenormand2009; @Vindenes2015; @Schreiber2015]. In this we follow @Levins1966 in
sacrificing realism for generality.

In this spirit, I provide concise, detailed and commented
code for simulating each of the models presented in Appendix A.
Simulations permit poking and prodding of empirically-minded
investigation unencumbered by either experimental design or mathematical formalism.
Most stochastic models are expressed in the BUGS language, which
may be more familiar to empirical readers than corresponding
mathematical formulas. This also permits both efficient simulation
and potential estimation of parameters given sample data using the
R package, NIMBLE [@deValpine2017]; allowing the deductive models
illustrated here to readily function as inductive models with
parameters inferred from time-series data. A copy of this appendix
is maintained at <https://github.com/cboettig/noise-phenomena>, and
bug reports, suggestions or help requests are welcome through the
issue tracker.  Appendix B provides more mathematical details
for interested readers.  With these in place, we are ready to begin
our tour through three major areas of noisy phenomena: Origins of
noise, emergent phenomena, and noise-driven inference.

>>>>

@@

@@

<!-- -->

The gap between specialists and non-specialists is particularly acute in the
ecological literature on stochastic processes.  This makes the potential benefit
for new synthesis all the greater, but also poses an obvious challenge in 
appealing to both audiences.  In recognition of this, I provide concise and
commented code for simulating each of the models presented in the supplement,
Appendix A, while more mathematical background and derivations are presented Appendix B.
Numerical simulations permit poking and prodding of empirically-minded
investigation unencumbered by either experimental design or mathematical formalism.
Most stochastic models are expressed in the BUGS language, which
may be more familiar to empirical readers than corresponding
mathematical formulas. This also permits both efficient simulation
and potential estimation of parameters given sample data using the
R package, NIMBLE [@deValpine2017]; allowing the deductive models
illustrated here to readily function as inductive models with
parameters inferred from time-series data. A copy of this appendix
is maintained at <https://github.com/cboettig/noise-phenomena>, and
bug reports, suggestions or help requests are welcome through the
issue tracker.

To emphasize the underlying trend in the changing roles in which we
see and understand noisy processes, I will also restrict my focus 
to relatively simple models primarily from population ecology context.
Simplicity not only makes examples (in equations and in code) more
tractable but also allows us to focus on aspects that are germane to many
contexts rather than unique to particular complexitites [@Levins1966; @Bartlett1960].
Nevertheless, that complexity matters -- few themes have been better 
emphasized in the theoretical literature [@Bjornstad2001].  Both
the foundational literature and recent research continue to echo the
theme of understanding the impact different real world complexities
have in stochastic dynamics, including the importance
of age and stage structure [e.g. @Cohen1979; @Tuljapurkar1989;
@Caswell2009; @Metcalf2015; @Vindenes2017b], spatial structure [e.g.
@Durrett1994; @Dieckmann2000; @Kerr2002; @Schreiber2010], Individual
heterogeneity [e.g. @Vindenes2015; @Hart2016] or the coexistence
of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007;
@Schreiber2017]; correlated [@Ripa; @Ripa1999; @Petchey, @Schreiber2010;
@Lee2017; @Spanio2017] or periodic [@Coulson2001; @Bjornstad2001;
@Keeling2001] structure in environmental noise, or the interaction
of ecological and evolutionary processes [@Saether1997; @Ozgul2009;
@Lenormand2009; @Vindenes2015; @Schreiber2015].  As such, we will 
rely on both textbooks and recent reviews to provide a proper treatment
of these issues, and focus on broader trends that may be less apparent
to non-specialists.

This review is structured into three sections: Origins of noise, emergent
phenomena, and noise-driven inference.  The first section lays the conceptual
groundwork we will need, while also highlighting a shift to more and more
mechanistically rooted descriptions of noise.  We will see where the common
formulation of 'deterministic skeleton plus noise term' comes from, how it 
is best justified, and when it breaks down.  The second section introduces
"noise the creator," showing examples of ecological phenomena generated by
stochasticity.  These examples will be familiar to most specialists but
illustrate a different way of thinking than held by most ecologists, where
noise is only a nuciance to be filtered or averaged out.  The third and
final section, "noise the informer," turns these examples back-to-front, 
asking what noise can tell us about a system: such as its underlying resilience
or stability, or the approach of a catastophic shift. Examples are fewer
here, and have largely yet to benefit from the introduction of either
the rigorous theorems or more complex models so plentiful in the previous
sections.  Yet the promise of prediction, of early warning signs before
tipping points, have spurred broad interest in such noise-based inference.
This review is a call to both deepen the connection to mechanism and better
formalize this thinking, but also look more broadly into other ways in which
noisy phenomena can help inform and predict underlying processes.
